---
title: "hw2"
author: "Jiahao Tian"
date: "2023-01-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 2

# Question 1

- Show $\sqrt{{{n_r}\over n}}$ is mle of $p_r$.


$$\begin{aligned}
n_r & = number\;of\;people\;with\;genotype\;rr\\
n_R & = number\;of\;people\;with\;genotype\;Rr\;and\; RR\\
p(_{RR})+p(_{Rr}) &=  p^2_R + 2p_Rp_r \\
p(_{rr}) & =p^2_{r} 
\end{aligned}$$



- Find likelihood and log-likelihood, then take a derivative set to equal 0.


$$\begin{aligned}
L(p_R,p_r) & = k(p^2_R + 2p_Rp_r)^{n_R}(p^2_r)^{n_r},\;k\;is\;constant\;here\\
\;\\
Becuase\;under\;HWE:\;\;\; n &= n_r + n_R\;\;\;\;\;p_r + p_R =1\\
\;\\
L(p_r) &= k(1-p^2_r)^{(n-n_r)}(p^2_r)^{n_r}\\
\;\\
log(L(p_r)) & \propto (n-n_r)log(1-p^2_r)+n_rlog(p^2_r)\\
\;\\
\triangledown log(L(p_r)) & \propto-{{(n-n_r)2p_r}\over 1-p^2_r}+{{2n_r}\over p_r}\\
\;\\
& Set \;to\;equal\; 0\\
\;\\
{{2n_r}\over p_r} & = {{2p_r(n-n_r)}\over 1-p^2_r}\\
\;\\
2n_r-2p^2_rn_r & = 2np^2_r-2p^2_rn_r\\
\;\\
\hat p_r & = \sqrt{{{n_r}\over n}}\\
\end{aligned}$$


- Check second derivative is less than 0.


$$\begin{aligned}
\triangledown^2 log(L(p_r)) & = -{{2(np^4_r +(n-3n_r)p^2_r+n_r)}\over p^2_r(p^2_r -1)^2} < 0 \\
\end{aligned}$$






# Question 5

- Observed data likelihood

$$\begin{aligned}
L^O(p_{AB}, p_{Ab}, p_{aB}, p_{ab}) & \propto (p_{AB}^2)^{n_{AABB}}(p_{Ab}p_{AB})^{n_{AABb}}(p_{Ab}^2)^{n_{AAbb}}\\
& (p_{AB}p_{aB})^{n_{AaBB}}(2p_{AB}p_{ab}+2p_{Ab}p_{aB})^{n_{AaBb}}(p_{Ab}p_{ab})^{n_{Aabb}}\\
& (p_{aB}^2)^{n_{aaBB}}(p_{aB}p_{ab})^{n_{aaBb}} (p_{ab}^2)^{n_{aabb}} \\
\end{aligned}$$


- Complete data likelihood

$$\begin{aligned}
L^C(p_{AB}, p_{Ab}, p_{aB}, p_{ab}) & \propto (p_{AB}^2)^{n_{AABB}}(p_{Ab}p_{AB})^{n_{AABb}}(p_{Ab}^2)^{n_{AAbb}}\\
& (p_{AB}p_{aB})^{n_{AaBB}}(2p_{AB}p_{ab})^{n_{AB|ab}}(2p_{Ab}p_{aB})^{n_{Aa|Bb}}(p_{Ab}p_{ab})^{n_{Aabb}}\\
& (p_{aB}^2)^{n_{aaBB}}(p_{aB}p_{ab})^{n_{aaBb}} (p_{ab}^2)^{n_{aabb}} 
\end{aligned}$$



- The two double heterozygous $n_{AaBb}$ are "missing data" due to ambiguity.

# First step, Expectation:

- Assume an initial value of $p_{AB},p_{Ab}, p_{aB}, p_{ab}$, and estimate the "missing data" as proportions of the total people of $n_{AaBb}$ of double heterozygous.


$$\begin{aligned}
Expected \;to\; see:\\
p(AB|ab) & = 2p_{AB}p_{ab}\\
p(Ab|aB) & = 2p_{Ab}p_{aB} \\
\;\;\;\;\;\;\; \\
E(n_{AB|ab}\mid p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m) & = \frac{p_{AB}^mp_{ab}^m}{p_{AB}^mp_{ab}^m + p_{Ab}^mp_{aB}^m}n_{AaBb}\\
E(n_{Ab|aB}\mid p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m) & = \frac{p_{Ab}^mp_{aB}^m}{p_{AB}^mp_{ab}^m + p_{Ab}^mp_{aB}^m}n_{AaBb}\\
{n_{AB|ab}}^m & = \frac{p_{AB}^mp_{ab}^m}{p_{AB}^mp_{ab}^m + p_{Ab}^mp_{aB}^m}n_{AaBb}\\
\end{aligned}$$


- So the number of people with genotype AB|ab from our "guess" is as above.
- Where "m" is the current step.



- Then need to construct a function Q and maximize function Q to get a new estimator .s.t $(P_{AB})^{m+1}$ of the parameter. Then use the new "guess" of estimator $(P_{AB})^{m+1}$ to create new guess at what the value of $n_{AaBb}$ are and then give the value of $n_{AaBb}$ to construct a new function Q and maximize the new function Q to obtain the results.
- Function $Q = E_{n_{AB|ab}, n_{Aa|Bb} | p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m}[log(complete\;data \;likelihood)]$


- Complete data log-likelihood

$$\begin{aligned}
log(L^C(p_{AB}, p_{Ab}, p_{aB}, p_{ab})) & \propto \left[ 2n_{AABB} + n_{AABb} + n_{AaBB} + n_{AB|ab}\right]\ln p_{AB} + \\
&\left[ n_{AABb} + 2n_{AAbb} + n_{Aa|Bb} + n_{Aabb}\right]\ln p_{Ab} +\\
&\left[ n_{AaBB} + n_{Aa|Bb} + 2n_{aaBB} + n_{aaBb}\right]\ln p_{aB} +\\
&\left[ n_{AB|ab} + n_{Aabb} + n_{aaBb} + 2n_{aabb}\right]\ln p_{ab}
\end{aligned}$$


- Function Q

$$\begin{aligned}
Q & = E_{n_{AaBb}, n_{Aa|Bb}| p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m}[log(L^C(p_{AB}, p_{Ab}, p_{aB}, p_{ab}))]\\
& \propto \left[ 2n_{AABB} + n_{AABb} + n_{AaBB} + E(n_{AB|ab}\mid p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m) \right]\ln p_{AB} + \\
&  \left[ n_{AABb} + 2n_{AAbb} + E(n_{Aa|Bb}\mid p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m) + n_{Aabb}\right]\ln p_{Ab} +\\
&  \left[ n_{AaBB} + E(n_{Aa|Bb}\mid p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m) + 2n_{aaBB} + n_{aaBb}\right]\ln p_{aB} +\\
&   \left[ E(n_{AB|ab}\mid p_{AB}^m, p_{Ab}^m, p_{aB}^m, p_{ab}^m) + n_{Aabb} + n_{aaBb} + 2n_{aabb}\right]\ln p_{ab}
\end{aligned}$$


# Second step, Maximization:

- An initial estimate $p_{AB}^m$ is put into the right hand side to give an updated estimated $p_{AB}^{m+1}$ on the left hand side.


$$\begin{aligned}
p_{AB}^{m+1} & = {{2n_{AABB}+n_{AABb}+n_{AaBB}+n_{AB|ab}^{m}}\over 2n}\\
p_{AB}^{m+1}2n & = 2n_{AABB}+n_{AABb}+n_{AaBB}+\frac{p_{AB}^mp_{ab}^m}{p_{AB}^mp_{ab}^m + p_{Ab}^mp_{aB}^m}n_{AaBb}\\
\end{aligned}$$


- Implement this EM algorithm on the mosquito data.

$$\begin{aligned}
& From\;the\;table\;'n=40'\\
& n_{AA} = 25,\;n_{Aa} = 16,\;n_{aa}=0 \\
& n_{BB} = 27,\;n_{Bb} = 13,\;n_{bb}=0\\
& \;\;\; \\
& Then\;we\;can\;get\;'2n=80' \\
& n_{A} = 64,\;n_{a}=16\\
& n_{B} = 67,\; n_{b}=13\\
&\;\;\;\; \\
& p_{A} = 0.8,\;p_{B}=0.8375,\;p_{a} = 0.2,\;p_{b} = 0.1625\\
&\;\;\\
&Our\;initial\;starting\;point:p_{AB}^m = p_{A}*p_{B},\;so\;on\\
&each\;iteration\;will\;update\;n_{AaBb}\\
\end{aligned}$$


- Plug above information back to the equation of $p_{AB}^{m+1}$.

$$\begin{aligned}
p_{AB}^{m+1} & = {{38+5+8+n_{AB|ab}^{m}}\over 2n}\\
p_{AB}^{m+1}2n & = 51+\frac{p_{AB}^mp_{ab}^m}{p_{AB}^mp_{ab}^m + p_{Ab}^mp_{aB}^m}n_{AaBb}\\
\end{aligned}$$


```{r}
geno = function(pAB0, pab0, pAb0, paB0, n_iter) {
  
nAABB = 19
nAABb = 5
nAaBB = 8
nAaBb = 8
nAAbb = 0
nAabb = 0
naabb = 0
naaBb = 0
naaBB = 0
n = nAABB + nAABb + nAaBB + nAaBb

pAB = rep(0, n_iter)
pab = rep(0, n_iter)
pAb = rep(0, n_iter)
paB = rep(0, n_iter)

pAB[1] = pAB0
pab[1] = pab0
pAb[1] = pAb0
paB[1] = paB0

for(i in 1:n_iter) {
  nABabm = ((pAB[i] * pab[i]) / (pAB[i] * pab[i] + pAb[i] * paB[i])) * nAaBb
  nAbaBm = ((pAb[i] * paB[i]) / (pAb[i] * paB[i] + pAB[i] * pab[i])) * nAaBb
  
  pAB[i+1] = (2 * nAABB + nAABb + nAaBB + nABabm) / (n * 2)
  pAb[i+1] = (nAABb + 2 * nAAbb + nAbaBm + nAabb) / (n * 2)
  paB[i+1] = (nAaBB + nAbaBm + 2 * naaBB + naaBb) / (n * 2)
  pab[i+1] = (nABabm + nAabb + naaBb + 2 * naabb) / (n * 2)
}

list(pAB = pAB, pab = pab, pAb = pAb, paB = paB)

}
```



```{r}
set.seed(22)
pAB0 = 0.67
pab0 = 0.33
pAb0 = 0.134
paB0 = 0.1675
geno(pAB0, pab0, pAb0, paB0, 5)
```


# Chapter 4

# Question 1

- $H_0$: HWE holds in the population.
- $H_a$: HWE does not hold in the population.


$$\begin{aligned}
Phenotype&\;\;:MM\;\;\;MN\;\;\;NN \\
Genotype&\;\;:M/M\;\;M/N\;\;N/N \\
(Observed)\;Number&\;\;:119\;\;\;\;\;\;\;76\;\;\;\;\;\;\;13\\
\;\;\\
\hat p_M &= {{2*119+76}\over 2*208}=0.755\\
\;\;\\
\hat p_N &= {{2*13+76}\over 2*208} = 0.245\\
\;\;\\
& Expected\;value\\
n_{MM} & = 208 * 0.755^2 = 118.57 \\
n_{NM} & = 208 * 2 * 0.755*0.245 = 76.95\\
n_{NN} & = 208*0.245^2 = 12.49\\
\;\;\\
\chi^2 & = \sum{{(O-E)^2}\over E}\\
& = {{(119-118.57)^2\over 118.57}}+{{(76-76.95)^2\over 76.95}}+{{(13-12.49)^2}\over 12.49}\\
& = 0.0341
\end{aligned}$$


- With DF = 1, and at $\alpha= 0.05$, 0.034 < $\chi^2_{1,0.05}=3.841$. We do not reject $H_0$. Thus the population is in HWE.




# Question 2


- $H_0$: HWE holds in the population.
- $H_a$: HWE does not hold in the population.


$$\begin{aligned}
Female\;:\\
&t/t = 63\;\;t/y=55\;\;y/y=12\\
Male\;:\\
&t = 74\;\;\;\;\;y=38\\
\end{aligned}$$


- Phenotypic frequencies


$$\begin{aligned}
Female\;:\\
&\hat q_{fy} = {{12}\over 130} = 0.092;\;\;\hat q_{ft} = {{63}\over 130} =0.485;\;\;\hat q_{fty} = {{55}\over 130} = 0.423\\
\;\\
Male\;:\\
&\hat q_{my} = {{38}\over 112}=0.339;\;\;\hat q_{mt} = {{74}\over 112} = 0.661\\
\end{aligned}$$


- Gene counting (both male and female)


$$\begin{aligned}
\hat p_{t} & = {{2*63+55+74}\over (63+55+12)*2+74+38} = 0.685\\
\;\;\\
\hat p_{y} & = {{55+2*12+38}\over (63+55+12)*2+74+38} = 0.315\\
\end{aligned}$$


- LRT

- nmt = number of male is genotype t, nft = number of female is genotype t/t, etc.

$$\begin{aligned}
2log{{L(\hat q)}\over L(\hat p)} & = 2log\bigg({{\hat q_{fy}^{nfy}*\hat q_{ft}^{nft}*\hat q_{fty}^{nfty}*\hat q_{mt}^{nmt} *\hat q_{my}^{nmy}\over (\hat p_t^2)^{nft}*(\hat p_y^2)^{nfy}* (2\hat p_t\hat p_y)^{nfty}*(\hat p_t)^{nmt}*(\hat p_y)^{nmy} }} \bigg)\\
& = 2log(1.254)\\
& = 0.1966\\
\end{aligned}$$


- With DF = 1, and at $\alpha= 0.05$, 0.1966 < $\chi^2_{1,0.05}=3.841$. We do not reject $H_0$. Thus the population is in HWE.





